---
title: "Mini-Project - Global Temperatures"
author: "Jordan Menter & Lizzie Kumar"
date: "June 7, 2016"
output: html_document
---

### Introduction

In this report, we analyze trends in average monthly global temperature from 1881 to 2012. Our [data](http://data.giss.nasa.gov/gistemp/tabledata_v3/GLB.Ts+dSST.txt) comes from the NASA Goddard Institute for Space Studies. Note that this data is measured in 0.01 degrees Celsius - for our analyses, we convert it to absolute temperature in Fahrenheit. 

```{r temp_convert, echo = FALSE}
# Load data
load(url('http://www.stolaf.edu/people/olaf/cs125/project1.RData'))

# Convert to absolute temperature in F
# Add 14, then mult. by 1.8 then + 32
faren_data <- (tData/100 + 14)*1.8 + 32

```

### Monthly Mean Global Temperatures

```{r global_temps, echo = FALSE}
# Separate vectors for each month - function
data_matrix <- matrix(data = faren_data, nrow = length(faren_data)/12, ncol = 12, byrow = TRUE)

# Set column names
colnames(data_matrix) <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Mean global temperatures
month_means <- colMeans(data_matrix)

# Lowest and highest global mean temperatures for each month
month_max <- apply(data_matrix, 2, max)
month_min <- apply(data_matrix, 2, min)

# When did these temperatures occur
month_max_index <- apply(data_matrix, 2, which.max)
month_min_index <- apply(data_matrix, 2, which.min)

# Get dates
# month_max_dates <- apply(month_max_index, 2, get_dates[2])
```

We begin by finding the mean global temperature for each month. This data is displayed in the following table:

`r month_means`

We also find in which year the highest average global temperature occurs for each month, displayed in the following table. Note that the values are the months since January 1881.

`r month_max_index`

We complete the same analysis for the lowest average global temperature of each month, shown below.

`r month_min_index`

Notice that the dates at which the highest average global temperatures occur are consistently more recent than those at which the lowest average global temperatures occur. 

```{r index_functions, echo = FALSE}
# Get index for given month and year
get_index <- function(month, year) {
  return((year - 1881)*12 + month)
}

# Get month and year for given index
get_dates <- function(index) {
  month <- index %% 12
  if (month == 0) month <- 12
  year <- (index - month)/12 + 1881
  return(c(month, year))
}
```

### Visualizing Monthly Average Global Temperatures

``` {r month_plots, echo = FALSE}
# set year vector
all_years <- 1881:2012

# set plotting grid
par(mfrow=c(3,4))

# plot each month
for (i in 1:12) {
  plot(all_years, data_matrix[,i], xlab = "Year", ylab = "Temperature",
       main = colnames(data_matrix)[i], type = "l")
}
```

We see that each month has a consistent warming trend over the course of the years under consideration. 

### Visualizing Average Annual Temperatures, 1881 - 2012

```{r annual_plot, echo = FALSE}
annual_temp <- rowMeans(data_matrix)
plot(all_years, annual_temp, xlab = "Year", ylab = "Temperature",
     main = "Average Annual Temperature", type = "l")
```

As in the previous plots, we also see a consistent warming trend in the average annual temperatures over all years. The lowest average temperature occured in `r which.min(annual_temp) + 1880`, and the highest average temperature occured in `r which.max(annual_temp) + 1880`.

### Monthly Average Temperatures in the 2000s

```{r 2000s_plot, echo = FALSE}
# identify subset of original vector corresponding to 2000s
jan2000 <- get_index(1, 2000)
dec2012 <- get_index(12, 2012)

plot(jan2000:dec2012, faren_data[jan2000:dec2012], type='l', xlab = "Months Since 1881", ylab = "Temperature", main = "2000s Average Monthly Temperature")
```

When considering only the years from 2000 to 2012, it is difficult to discern a consistent warming or cooling trend. The data appears rather noisy.

### Monthly Average Temperatures in the 1990s

```{r 1990s_plot, echo = FALSE}
# Identify subset of original vector corresponding to 1990s
jan1990 <- get_index(1, 1990)
dec1999 <- get_index(12, 1999)

plot(jan1990:dec1999, faren_data[jan1990:dec1999], type='l', xlab = "Months Since 1881", ylab = "Temperature", main = "1990s Average Monthly Temperature")

# Which months had the highest and lowest temperatures?
nineties_subset <- faren_data[jan1990:dec1999]
# Find maximum
max_index <- which.max(nineties_subset) + jan1990 - 1
max_dates <- get_dates(max_index)
# Find minimum
min_index <- which.min(nineties_subset) + jan1990 - 1
min_dates <- get_dates(min_index)
```

### Monthly Average Temperatures in the 1890s

```{r 1890s_plot, echo = FALSE}
# Identify subset of original vector corresponding to 1890s
jan1890 <- get_index(1, 1890)
dec1899 <- get_index(12, 1899)

plot(jan1890:dec1899, faren_data[jan1890:dec1899], type='l', xlab = "Months Since 1881", ylab = "Temperature", main = "1890s Average Monthly Temperature")

# Mean temperature of this decade
# mean(faren_data[jan1890:dec1899])
```

The mean temperature of this decade is `r mean(faren_data[jan1890:dec1899])`.

### Observing Warming Trends

```{r 44_years_plots, echo = FALSE}
# First 44 years
jan1881 <- get_index(1, 1881)
dec1924 <- get_index(12, 1924)
first_44 <- faren_data[jan1881:dec1924]
plot(jan1881:dec1924, first_44, type = 'l', xlab = "Months Since 1881", ylab = "Temperature", main = "Monthly Avg. Temp. - First 44 Years")
# Average Temperature
# mean(faren_data[jan1881:dec1924])

# Second 44 years
jan1925 <- get_index(1, 1925)
dec1968 <- get_index(12, 1968)
second_44 <- faren_data[jan1925:dec1968]
plot(jan1925:dec1968, second_44, type = 'l', xlab = "Months Since 1881", ylab = "Temperature", main = "Monthly Avg. Temp. - Second 44 Years")
# Average Temperature
# mean(faren_data[jan1925:dec1968])

# last 44 years
jan1969 <- get_index(1, 1969)
dec2012 <- get_index(12, 2012)
last_44 <- faren_data[jan1969:dec2012]
plot(jan1969:dec2012, last_44, type = 'l', xlab = "Months Since 1881", ylab = "Temperature", main = "Monthly Avg. Temp. - Last 44 Years")
# Average Temperature
# mean(faren_data[jan1969:dec2012])

```

The mean temperature of the first 44 years is `r mean(faren_data[jan1881:dec1924])`. 
Of the three periods considered here, it is easiest to observe a consistent warming trend in the last 44 years.